Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory

### 2022-11-09 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@

@@

@@

In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss. A \textit{de novo} gene emerges from a non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation. Both transcription and translation are complex processes involving many biomolecular complexes that work in concert. Here, we focus on the minimal requirements for these processes to occur. Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence. [here a rationale for this choice]. For translation, we only require the gene to have an open reading frame (ORF).

Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost via mutagenesis. These probability models are essentially described by two kinds of probability. First is the probability of finding a DNA sequence feature (such as TATA box or an ORF). This probability depends on the nucleotide distribution given by the GC content. The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another. Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.

For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes. We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}. We used mutation bias data from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}. shows the values of different mutation rates (\autoref{mutbias}).

@@

Overall our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone. A significant deviation from this null hypothesis may be interpreted as evidence in support of natural selection.

@@

\begin{figure}[!b]
\hrule
\vspace{1ex}
\centering
\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e. $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF. \autoref{genegaineq} describes total gene gain probability. Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$. Using the conditional probability of gene gain does not change these results.}
\label{gainlossprob}
\end{figure}

@@

@@

\begin{figure}[!b]
\hrule
\vspace{1ex}
\centering
%\begin{subfigure}{0.48\textwidth}
%\centering
%\caption{One-step}
\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
%\end{subfigure}\hfill%
%\begin{subfigure}{0.48\textwidth}
%\centering
%\caption{Two-step}
%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
%\end{subfigure}
\caption{\textit{De novo} genes with shorter ORFs ($<$ 42 codons) preferentially emerge from existing ORFs (ORF-first) whereas \textit{de novo} genes with longer ORFs preferentially emerge from existing RNAs (RNA-first). The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}). A positive value suggests a ORF-first trajectory and a negative value supports RNA-first trajectory. Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
\label{whoisfirst}
\end{figure}

For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge. That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain. Thus it may appear that the order of the occurrence of these two events does not matter. Gene gain is much more likely when one of the two features already exist (\autoref{genegaineq}) and hence there are two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first). To test the hypothesis that the order of feature gain events influences \textit{de novo} emergence, we first asked how much more likely it is to lose transcription before gaining an ORF. Specifically, we calculated the fold difference between the probability values of transcription loss and ORF gain. We found that transcription loss is $10^3$ to $10^{10}$ times higher than ORF gain, depending on the ORF length. This suggests that a mutation has a higher chance of disrupting an existing untranscribed ORF than cause a gain of transcription features. Next, we calculated the fold difference between transcription gain and ORF loss probability values, and found that ORF loss is $10^4$ to $10^5$ times more probable than transcription gain, suggesting that in a transcribed DNA region, a new mutation will more likely disrupt the transcription features than cause a gain of ORF. More generally features are more easily lost due to mutations than they are gained. However, this effect of mutations is not uniform for the two evolutionary trajectories and for different length of ORFs. For example, gain of an 40 codon long ORF in a transcribed region of DNA is $9.2\times10^3$ less likely than the transcription being lost. However, if the ORF already exists, then it is $2.7\times10^4$ times more likely to be lost before transcription emerges. This suggests that the two trajectories of \textit{de novo} emergence have different dynamics.

To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities. First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF. This probability denotes the trajectory where ORF emerges first.

@@

The second probability that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription.

@@

@@

Overall, our analysis suggests that \textit{de novo} genes with long ORFs are more likely arise from DNA regions that are already transcribed (for example, non-coding RNAs).

@@

We reframed our question to make it more relatable to commonly observed findings from genomics experiments. For example, a hypothetical experiment involves DNA sequencing of a genomic locus in different related species. This experiment shows that transcription features are present in only one species, whereas the ORF is present in all species. Two evolutionary events could describe this observation. First possibility is that a \textit{de novo} gene existed in an ancestor and transcription was lost in multiple species. For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$. (Note that $P_\textit{RNA-loss}$ is a conditional probability dependent on the presence of RNA).

@@

We answer the long standing question of what is the sequence of \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.

@@

As the famous saying goes, ``all models are wrong but some are useful'' \citep{GEBox}. This is true for our models too. They are not accurate but they can be useful for making predictions that can lead a more focused experimental validation. These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution. Therefore our work opens up an opportunity for theoretical biologists and computational biologists to develop models that are accurately describe their system of interest.

@@

\bibliographystyle{plainnat}
\bibliography{refs}

>>>>

\renewenvironment{abstract}{%
  \normalsize
  \textbf{\large\abstractname\\[1em]} % with a normal space
}

\title{Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory}
\author{Bharat Ravi Iyengar, Erich Bornberg-Bauer}
\date{\small Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany}

@@

\maketitle

\begin{abstract}
New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process known as \textit{de novo} gene emergence. To synthesize a protein, DNA must be transcribed as well as translated, both of which need some DNA sequence features. Stable transcription requires promoters and polydenylation signal, and translation requires at least an open reading frame (ORF). We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly do genes emerge and are lost. We also investigate whether different DNA features evolve in a sequential manner, and if sequence composition is biased by mutation rate. We find out that genes are lost much more rapidly than they emerge, and genes with long ORFs preferentially arise in regions that are already transcribed. Our study not only answers some fundamental question on the topic, but also provides a modeling framework for future studies.
\end{abstract}

@@

Organisms evolve new traits by expanding their functional genome. Evolution of new genes is one of the ways by which new traits can emerge. The definition of a gene is complicated, and has been changing constantly \citep{genepostencode}. We use the following working definition of a gene: a gene is a region of the genome that gives rise to a functional product, that is an RNA or a protein. The next big challenge is to define what a function is, and how to classify gene functions. One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions. Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins (mRNA) or participate in protein synthesis (rRNA, tRNA), are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.

In this study we focus on the evolution of genes that encode new proteins. New protein coding genes frequently arise via duplication of existing protein coding genes. These duplicated genes then genetically and functionally diverge from their parent genes by accumulating mutations \citep{Long2003,IAD}. Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein. This phenomenon is called \textit{de novo} gene emergence \citep{EBB-F1000,vanOss}. \textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{vanOss}. To express a protein, a genomic sequence should be transcribed as well as translated. Therefore, the gene needs sequence features that enable both these processes.

The primary requirement for transcription is recruitment of RNA polymerase at the gene's promoter \citep{Promoters}. A part of the promoter, called the core promoter is the region that determines the start of transcription \citep{corepromoters}. The second requirement for a productive transcription is its termination at appropriate sites. Even though transcription is widespread in eukaryotic genomes, a phenomenon known as pervasive transcription, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}. Most eukaryotic mRNAs and long non-coding RNAs, are polyadenylated at their 3' termini, which makes them resistant to degradation. Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals. In prokaryotes, end of transcription is determined by sequences known as terminators \citep{ProkTerm}. Thus the synthesis of a stable RNA, requires promoters, as well as termination/polyadenylation signal sequences.

The second major requirement for protein expression is translation of the transcribed mRNA. The most fundamental requirement for translation is the presence of an open reading frame (ORF). Usually, an mRNA needs additional features to initiate protein synthesis. These features include ribosome binding sites in prokaryotes \citep{RBS}, and Kozak consensus sequences \citep{kozak} in eukaryotes.

Although gain of transcription and translation features guarantees \textit{de novo} birth of a protein coding gene, it does not ensure that the gene would persist in the genome for many generations. The newly born gene can lose the features as easily as it gained them unless it is fixed in the genome via evolutionary selection. If the protein synthesized by the \textit{de novo} gene provides some fitness advantage to the host organism then it will undergo positive selection. Simulataneously, the protein should have low toxicity and cost of synthesis, to survive purifying selection. A common mechanism by which protein mediated toxicity occurs is by misfolding and aggregation of proteins \citep{misfold1,misfold2}.

In this study we develop probabilistic models of \textit{de novo} gene emergence. We focus on \textit{de novo} genes emerging from intergenic sequences, also known as ``protogenes'' \citep{protogenes}. We model \textit{de novo} gene emergence as a two step process. Specifically, the protogene can first gain transcription features and produce an untranslated RNA, which in turn acquires translation features subsequently. Alternatively, the protogene can first gain translational features, and later gain transcription. Our models are based on neutral evolution where natural selection does not affect the rate of gene emergence, or loss. Using these models, we predict that \textit{de novo} gene emergence has a preferred evolutionary trajectory, even under the assumption of neutrality.

@@

In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss. A \textit{de novo} gene emerges from a non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation. Both transcription and translation are complex processes involving many biomolecular complexes that work in concert. Here, we focus on the minimal requirements for these processes to occur. Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence. We chose these two promoter motifs because they are the most abundant type of core promoters across diverse eukaryotes \citep{Promoters,corepromoters}. We defined the poly-A signal as a transcription requirement because it is crucial for transcript stability and its export to cytoplasm\citep{polyAterm,polyAexp}. For translation, we only require the gene to have an open reading frame (ORF).

Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost via mutagenesis. These probability models are essentially described by two kinds of probability. First is the probability of finding a DNA sequence feature (such as TATA box or an ORF; \autoref{methprob}). This probability depends on the nucleotide distribution given by the GC content. The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another (\autoref{methgain}). Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.

For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes. We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}. We used mutation bias data (\autoref{mutbias}) from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.

@@

Overall our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.

@@

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e. $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF. \autoref{genegaineq} describes total gene gain probability. Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$. Using the conditional probability of gene gain does not change these results.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

@@

For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge. That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain. Thus it may appear that the order of the occurrence of these two events does not matter. However, gene emergence is much more likely when one of the two features already exist (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first). To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities. First is the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF. This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$; \autoref{orffirsteq1}).

@@

The second probability ($P_\textit{RNA-first}$; \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.

@@

\begin{figure}[!t]
\centering
%\begin{subfigure}{0.48\textwidth}
%\centering
%\caption{One-step}
\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
%\end{subfigure}\hfill%
%\begin{subfigure}{0.48\textwidth}
%\centering
%\caption{Two-step}
%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
%\end{subfigure}
\caption{\textit{De novo} genes with shorter ORFs ($<$ 42 codons) preferentially emerge from existing ORFs (ORF-first) whereas \textit{de novo} genes with longer ORFs preferentially emerge from existing RNAs (RNA-first). The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}). A positive value suggests a ORF-first trajectory and a negative value supports RNA-first trajectory. Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}

@@

This preference of \textit{de novo} gene emergence trajectory exists, because the rates of feature gain and the rates of feature loss are different. For example, ORFs are very likely to be lost before transcription emerges. Specifically, ORF loss is $\sim10^4$ to $\sim10^5$ times more probable than transcription gain, such that this difference increases with the ORF length. The converse is also true, that is transcription loss is more probable than ORF gain. However, this difference is not the same as that between ORF loss and transcription gain. Transcription loss is $\sim10^3$ to $\sim10^{10}$ times more likely than ORF gain, such that this difference is larger for longer ORFs. Thus, long ORFs are more likely to be lost before transcription emerges, than short ORFs. Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.

@@

We reframed our question to make it more relatable to commonly observed findings from genomics experiments. For example, a hypothetical experiment involves DNA sequencing of a genomic locus in different related species. This experiment shows that transcription features are present in only one species, whereas the ORF is present in all species. Two evolutionary events could describe this observation. First possibility is that a \textit{de novo} gene existed in an ancestor and transcription was lost in multiple species. For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$. (Note that $P_\textit{RNA-loss}$ is a conditional probability, that is dependent on the presence of transcription).

@@

We answer the long standing question of what is the sequence of \textit{de novo} emergence. That is, whether ORF emerges first or transcription emerges first. We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus. Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs, but this evolutionary process may be constrained by a different set of factors. Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the \textit{de novo} gene. For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory. On the other hand, the selection would have no effect on an ORF that is not transcribed. Thus positive selection will intensify the preference for RNA-first trajectory.

@@

A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}. This is true for our models too. They are not accurate but they can be useful for making predictions that can lead a more focused experimental validation. These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution. Therefore our work opens up an opportunity for theoretical biologists and computational biologists to develop models that are accurately describe their system of interest.

@@

\bibliographystyle{plainnat}

\small
\bibliography{refs}
